---
title: "Intro to R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## How does R work?

R is a statistical programming language. This means that we can write code in R that a computer reads and interprets to perform specific tasks.

We'll be using RStudio as our way to interact with R code. The panel labeled `Console` is where we submit code to R and read the output. 

Click on the icon in the upper left window of RStudio of a white square with a green plus sign, then select `R Script`. This will open a new panel for typing text. Save this file someplace on your computer you'll be able to find it later, and call it `Practice.R`. 

Try typing the following basic math in the text window. You can send this line of code to the console for R to run by holding down the `Control` key and pressing `Return`. 

```{r math}
3 + 3 
```
In the example above, the code entered appears in the gray box, and the result appears below it.

We can include notes to ourselves that describe what the code does by adding a hash sign (or pound sign, or hash tag) in front of the notes on the same line. This is called commenting code:

```{r comment}
# this is a code comment, which is useful for describing the purpose of the code that follows

# performing basic math
3 + 3 
3 + 4 # you can include comments on the same line as code
3+4 # spaces on either side of the plus sign aren't necessary, but it makes your code easier to read
```

## Using functions

```{r functions}
# functions are used to perform specific tasks


# round a number
round(3.1415)

# change the number of decimal places
round(3.1415, digits = 2)

```

## Assigning objects

```{r}
# assigning value to an object
weight_kg <- 55 # <- is an assignment operator
# recall object
weight_kg
# perform an operation on an object
2.2 * weight_kg
# assign a new value to object
weight_kg <- 57.5
# create new object from operation
weight_lb <- 2.2 * weight_kg
weight_kg <- 100
```

## Tidyverse

Tidyverse is the name of a set of packages (software) that allow us to do extra cool stuff with data in R.

```{r tidyverse, eval=FALSE}
# install tidyverse
install.packages("tidyverse")
# load package
library(tidyverse)
```

## Data visualization

```{r}
# access data (available with packages)
data(mtcars)
head(df)
?mtcars
```

```{r}
# scatterplot of miles per gallon and weight
qplot(mpg, wt, data=mtcars)
```

```{r}
# color points by number of cylinders, add smoothing lines
qplot(mpg, wt, data = mtcars, color = factor(cyl),
      geom=c("point", "smooth"))
```

```{r}
# boxplot of mpg by transmission
qplot(factor(am), mpg, data = mtcars,
      geom="boxplot")
```